Higher-Order unification for free!

### 2024-04-12 paper/main.tex
sec 2.3: corrected notation
sec 2.3: notation

\subsection{The problem: logic-program simulation}
% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
\foUnifPb is a list of pair of terms, each pair correspond to a unification
problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
elements extracted from $P_p$).
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each \Fo{} term in \foUnifPb into a \Ho{} term $t$.
We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} \}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:

>>>>

\subsection{The problem: logic-program simulation}
% \todo{dire che P e T sono liste di coppie, dire che P e T vengono anche usati come insiemi}
We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} of length $\mathcal{N}$.
\foUnifPb is a list of pair of terms, each pair correspond to a unification
problem. At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ (the small $l$ and $r$ represent respectively the left and right 
elements extracted from $P_p$).
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[4]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho'}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each pair of \Fo{} terms in \foUnifPb into a pair of \Ho{} terms $t$.
We write this translation $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the pair $(\hoUnifPb_{p_l}, \hoUnifPb_{p_r})$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:
